Human capital mediates natural selection in contemporary humans

### 2022-01-03 why-natural-selection.Rmd
Small tweaks

We also examine selection effects among respondents' parents, using information
on respondents' number of siblings. Effect sizes of polygenic scores are highly
correlated across the two generations. Median-splitting respondents by year of
birth, we find little evidence of change in effect sizes among either the
parents' or the respondents' generation. We also check whether selection effects
vary by socio-economic status and AFLB in the parents' generation. We have no
information about parents' income, so we use the 1971 Townsend deprivation score
of respondents' birthplace as a proxy [@townsend1987deprivation]. Results show
the same pattern as for respondents. Effect sizes are larger and more often
significant in the most deprived areas (Appendix Figure
\@ref(fig:plot-siblings-townsend)). Effects are larger among younger parents,
and change sign when controlling for AFLB, among both mothers and fathers
(Appendix Figure \@ref(fig:plot-age-birth-parents)). Lastly, we check for a
"quantity-quality tradeoff" between parents' number of children and number of
grandchildren. We do not find any: in fact, the correlation between *number of
siblings* and *number of children* is positive ($\rho$ = 
`r sib_chn_cor$estimate`, $p < 2 \times 10^{-16}$).

@@

Balancing selection reduces variance in the trait under selection, 
while diversifying selection increases variance. To check for these,
we rerun equation \@ref(eq:regression), adding a quadratic term in $PGS_i$,
and using our age/qualification weights. Scores for hip circumference and drinks
per week show significant balancing selection ($p < 0.05/33$, negative
coefficient). Scores for educational attainment (EA2 and EA3) show significant
diversifying selection ($p < 0.05/33$, positive coefficient), which reduces the
strength of selection against educational attainment at very high levels of the 
PGS. Figure
\@ref(fig:plot-purifying) plots predicted number of children against polygenic
score from these regressions.

We also checked for balancing selection in the parents' generation, using
age/qualification weights multiplied by the inverse of number of siblings. 
Scores for EA2 and EA3 again show significant diversifying selection 
($p < 0.05/33$, positive coefficient). Other scores including hip circumference 
and drinks per week are not significant.

```{r plot-purifying, fig.align = "center", fig.cap = "Purifying/diversifying selection: predicted number of children by polygenic score.", fig.height = 5}

@@

stopifnot(sig == c("dpw_substance_use", "EA2_noUKB", "EA3_excl_23andMe_UK",
                     "hip_combined"))

@@

op <- par(mfrow = c(2,2))
for (score in sig) {
  coef <- sig_all_terms %>% filter(score_name == score) %>% pull(estimate)
  curve(
         coef[1] + coef[2] * x + coef[3] * x^2,
         xlim = c(-4, 4), 
         xlab = pretty_names(score),
         ylab = "Predicted N children"
       )
}
par(op)

>>>>

We also examine selection effects among respondents' parents, using information
on respondents' number of siblings. Effect sizes of polygenic scores are highly
correlated across the two generations. Median-splitting respondents by year of
birth, we find little evidence of change in effect sizes among either the
parents' or the respondents' generation. We also check whether selection effects
vary by socio-economic status and AFLB in the parents' generation. We have no
information about parents' income, so we use the 1971 Townsend deprivation score
of respondents' birthplace as a proxy [@townsend1987deprivation]. Results show
the same pattern as for respondents. Effect sizes are larger and more often
significant in the most deprived areas (Appendix Figure
\@ref(fig:plot-siblings-townsend)). Effects are larger among younger parents,
and change sign when controlling for AFLB, among both mothers and fathers
(Appendix Figures \@ref(fig:plot-age-birth-parents-cross), 
\@ref(fig:plot-age-birth-parents)). Lastly, we check for a
"quantity-quality tradeoff" between parents' number of children and number of
grandchildren. We do not find any: in fact, the correlation between *number of
siblings* and *number of children* is positive ($\rho$ = 
`r sib_chn_cor$estimate`, $p < 2 \times 10^{-16}$).

@@

Balancing selection reduces variance in the trait under selection, while
diversifying selection increases variance. To check for these, we rerun equation
\@ref(eq:regression), adding a quadratic term in $PGS_i$, and using our
age/qualification weights. Scores for hip circumference show significant
balancing selection ($p < 0.05/33$, negative coefficient on quadratic term).
Scores for educational attainment (EA2 and EA3) show significant diversifying
selection ($p < 0.05/33$, positive coefficient), which reduces the strength of
selection against educational attainment at very high levels of the PGS. Figure
\@ref(fig:plot-purifying) plots predicted number of children against polygenic
score from these regressions.

We also checked for balancing selection in the parents' generation, using
age/qualification weights multiplied by the inverse of number of siblings.
Scores for EA2 and EA3 again show significant diversifying selection ($p <
0.05/33$, positive coefficient). Other scores including hip circumference were 
not significant.

```{r plot-purifying, fig.align = "center", fig.cap = "Purifying/diversifying selection: predicted number of children by polygenic score.", fig.height = 3}

@@

stopifnot(sig == c("EA2_noUKB", "EA3_excl_23andMe_UK", "hip_combined"))

@@

op <- par(mfrow = c(1, 3))
for (score in sig) {
  coef <- sig_all_terms %>% filter(score_name == score) %>% pull(estimate)
  curve(
         coef[1] + coef[2] * x + coef[3] * x^2,
         xlim = c(-4, 4), 
         xlab = pretty_names(score),
         ylab = "Predicted N children"
       )
}
par(op)
